---
title: "The minimally sufficient adjustment set (MSAS) algorithm"
author: "KFArnold"
date: "03/02/2022"
output: html_document
---

<style type="text/css">
  body{
  font-size: 14pt;
}
</style>

```{r setup, include = FALSE}
# Set default options
knitr::opts_chunk$set(echo = TRUE,
                      fig.width = 12, 
                      fig.height = 8)


#knitr::opts_knit$set(root.dir = getwd())

# Restore package library to last snapshot
packrat::restore()

# Load required libraries
library(tidyverse)
library(dagitty)
library(ggdag)
library(broom)

# Run source data processing script
source("./Script/Data_processing.R")
```

# The Context

What is the *total causal effect* of **limiting longterm illness** on **feeling optimistic about the future**?

## Drawing the Directed Acyclic Graph (DAG)

To calculate the desired causal effect, we must first specify all key variables which might be relevant for this context. Examples of key variables therefore include:

* Age
* Employment status
* Self-assessed health status
* Therapy

We can then construct a DAG which specifies the *hypothesised causal relationships* amongst the key variables identified.

```{r download-dag}
# Import published DAG from dagitty.net & create 'tidy' DAG 
dag <- downloadGraph(x = "dagitty.net/m3DXjao") 
tidy_dag <- dag %>%
  tidy_dagitty %>%
  node_status(as_factor = FALSE) %>%
  mutate(label = str_replace_all(string = name, 
                                 pattern = "_", 
                                 replacement = "\n")) 
# Plot DAG
tidy_dag %>%
  ggplot(aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_edges() +
  geom_dag_point(aes(color = status)) + 
  geom_dag_label_repel(aes(label = label)) +
  theme_dag(plot.background = element_rect(colour = "black", fill = NA))
```

## Calculating the Minimally Sufficient Adjustment Set (MSAS)

Using the `dagitty` package, we can calculate the minimally sufficient adjustment set (MSAS).

```{r calculate-msas}
# Calculate minimally sufficient adjustment set (MSAS)
msas <- adjustmentSets(x = dag, type = "minimal", effect = "total") %>% pluck(1)
```

According to the algorithm, adjusting for the following variables is sufficient for closing all backdoor paths between limiting longterm illness and feeling optimism about the future:

```{r print-msas, echo = FALSE}
msas
```

We can visualise these nodes on the DAG as common causes of the exposure and outcome.

```{r plot-msas, echo = FALSE}
# Plot DAG with MSAS
tidy_dag %>%
  mutate(status = ifelse(name %in% msas, "msas", status)) %>%
  ggplot(aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_edges() +
  geom_dag_point(aes(color = status)) + 
  geom_dag_label_repel(aes(label = label)) +
  theme_dag(plot.background = element_rect(colour = "black", fill = NA))
```

## Estimating the Causal Effect

To estimate the causal effect of limiting longterm illness on feeling optimistic about the future, we use data collected by the Health Survey for England (HSE) in 2019:

```{r data-peek, echo = FALSE}
data
```

We fit a logistic regression model of `Optimism ~ LLTI`, adjusted for all variables in the MSAS.

```{r fit-glm}
# Fit GLM for outcome ~ exposure, adjusted for MSAS & Survey_qtr
mod <- glm(formula = Optimism ~ LLTI + Age + Sex + Ethnicity + Education + BMI + Survey_qtr,
           family = "quasibinomial",
           weights = Sample_weight,
           data = data)
```

```{r extract-coefficient, echo = FALSE}
# Create tidy summaries of fitted model:
# (1) all coefficients except intercept
full_tidy_summary <- tidy(mod, conf.int = TRUE, exponentiate = TRUE) %>%
  select(term, estimate, conf.low, conf.high, p.value) %>%
  filter(!str_detect(term, "Intercept"))
# (2) exposure coefficient only
tidy_summary <- full_tidy_summary %>%
  filter(str_detect(term, "LLTI"))

# Extract exposure coefficient from tidy summary
exp_coef <- tidy_summary %>% 
  select(estimate) %>% 
  pull %>% 
  round(digits = 3)
exp_conf_low <- tidy_summary %>% 
  select(conf.low) %>% 
  pull %>% 
  round(digits = 3)
exp_conf_high <- tidy_summary %>% 
  select(conf.high) %>% 
  pull %>% 
  round(digits = 3)
```

This model estimates that having a limiting longterm illness causes the odds of feeling optimistic about the future to decrease by **`r (1 - exp_coef)*100` percent** `r paste0("(95% CI: ", paste((1 - exp_conf_high)*100, (1 - exp_conf_low)*100, sep = ", " ), ")")`.

```{r tidy-summary, echo = FALSE}
tidy_summary
```

```{r plot-coefficients, echo = FALSE}
full_tidy_summary %>%
  ggplot(aes(x = term, y = estimate)) +
  theme_minimal() +
  theme(legend.position = "none",
        axis.title.x = element_blank(),
        axis.title.y = element_text(size = 14),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
        plot.background = element_rect(colour = "black", fill = NA)) +
  labs(y = "Odds") +
  geom_hline(yintercept = 1) +
  geom_point(aes(color = str_detect(term, "LLTI"), size = 2)) +
  geom_linerange(aes(x = term, ymin = conf.low, ymax = conf.high,
                     color = str_detect(term, "LLTI"), lwd = 1.5)) +
  scale_x_discrete(limits = full_tidy_summary %>% pull(term),
                   labels = function(x) str_wrap(x, width = 15)) +
  scale_color_manual(values = c("grey80", "dodgerblue"))
```

